---
title: "Classification Trees "
author: "Emanuel Tafani"
date: "2026-02-06"
output: html_document
---

## Introduction
This project is based on a medical context. Specifically, it focuses on the classification of breast tumors as benign (B) or malignant (M) based on morphological features extracted from Fine Needle Aspiration (FNA) images.
This work shows the importance of early diagnosis.

## Dataset description
The dataset used in this project is the German Credit Data from the UCI Machine Learning Repository, a well‑known credit scoring dataset that classifies individuals as good or bad credit risks based on financial and demographic features.

Ten real-valued features are computed for each cell nucleus:

	a) radius (mean of distances from center to points on the perimeter)
	b) texture (standard deviation of gray-scale values)
	c) perimeter
	d) area
	e) smoothness (local variation in radius lengths)
	f) compactness (perimeter^2 / area - 1.0)
	g) concavity (severity of concave portions of the contour)
	h) concave points (number of concave portions of the contour)
	i) symmetry 
	j) fractal dimension ("coastline approximation" - 1)

```{r}
# Libraries
library(tidyverse)
library(rpart)
library(rpart.plot)
library(randomForest)
library(gbm)
library(caret)

#Nomi delle colonne
features <- c(
  "radius", "texture", "perimeter", "area", "smoothness",
  "compactness", "concavity", "concave_points", "symmetry",
  "fractal_dimension"
)

col_names <- c(
  "id",
  "diagnosis",
  paste0(features, "_mean"),
  paste0(features, "_se"),
  paste0(features, "_worst")
)
col_names

bc <- read.csv("C:\\Users\\emanu\\Desktop\\Stat Learning\\data.csv", header = FALSE)
colnames(bc) <- col_names

bc$diagnosis <- factor(bc$diagnosis, levels = c("B", "M")) #variabile categorica, non numerica.

bc$id <- NULL #Rimozione id

str(bc)
summary(bc$diagnosis)


```


## Exploratory data analysis

```{r}
# 1) Distribuzione della classe target
table(bc$diagnosis)
table(bc$diagnosis)/dim(bc)[1]

# 2) Boxplot 
library(ggplot2)

# radius_mean vs diagnosis
ggplot(bc, aes(x = diagnosis, y = radius_mean, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Radius Mean by Diagnosis", x = "Diagnosis", y = "Radius Mean") +
  theme_minimal()

# texture_mean vs diagnosis
ggplot(bc, aes(x = diagnosis, y = texture_mean, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Texture Mean by Diagnosis", x = "Diagnosis", y = "Texture Mean") +
  theme_minimal()
```




## Classification tree

## Random forest

## Boosting

## Discussion and conclusions