---
title: "German Credit Risk Classification using Tree-Based Methods"
author: "Emanuel Tafani"
date: "2026-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. INTRODUCTION

The aim of this project is to classify customers as good or bad credit risks using the German Credit dataset present in the UCI Machine Learning Repository.
The dataset contains both numerical and categorical variables describing financial and personal characteristics of customers.

Tree-based methods are well suited for this type of data, as they can naturally handle both categorical and numerical predictors without requiring excessive preprocessing.

The following models are considered: single decision Tree, Bagging, Random Forest and Boosting.


2. DATA LOADING AND PREPROCESSING

```{r}
cr <- read.table("german.data", header = F)
colnames(cr)[colnames(cr) == "V21"] <- "Classe"

str(cr) # Variabili 
qualitative_vars <- c(1,3,4,6,7,9,10,12,14,15,17,19,20,21)
cr[qualitative_vars] <- lapply(cr[qualitative_vars], as.factor)
str(cr) # Variabili qualitative trasformate in "Factor"
summary(cr$Classe)  # 1=GOOD, 2=BAD

cr$V7 <- ordered(cr$V7,
                     levels = c("A71","A72","A73","A74","A75"))
#Attribute 7:  (qualitative)
#Present employment since
#A71 : unemployed
#A72 :       ... < 1 year
#A73 : 1  <= ... < 4 years  
#A74 : 4  <= ... < 7 years
#A75 :       .. >= 7 years

cr$V17 <- ordered(cr$V17,
                      levels = c("A171","A172","A173","A174"))
#Attribute 17: (qualitative)
#Job
#A171 : unemployed/ unskilled  - non-resident
#A172 : unskilled - resident
#A173 : skilled employee / official
#A174 : management/ self-employed/
#  highly qualified employee/ officer

str(cr) # Le due variabili sono ora "Ordered Factor"
summary(cr$Classe) 

```

Qualitative variables are converted into factors. Some of these have an ordinal nature so have been defined as ordered factors.


3. TRAIN-TEST SPLIT

The dataset is split into a training set (70%) and a test set (30%).

```{r}
set.seed(2)

n_cr <- nrow(cr)
train_idx_cr <- sample(1:n_cr, size = floor(0.7 * n_cr))

cr_train <- cr[train_idx_cr, ]
cr_test  <- cr[-train_idx_cr, ]

y_test_cr <- cr_test$Classe

summary(cr_train$Classe)
summary(cr_test$Classe)

```


4. DECISION TREE

First I implement a single classification tree on the training data and evaluate its performance on the test data.

```{r}
set.seed(2)
library(tree)    # Classification and regression trees
tree_cr <- tree(Classe ~ .,
                     data = cr_train)
summary(tree_cr) # Numero di nodi terminali(Regioni/Foglie) = 14
                 # Classification error rate sul training set = 0.1957
                 # Variabili usate nella costruzione dell'albero
                        # V1(Status del CC), V2(durata prestito in mesi)
                        # V3(precedente storia creditizia),
                        # V4(scopo del prestito), V5(importo creditizio   richiesto),
                        # V6(risparmi o obbligazioni), V10(altri debitori/garanti),
                        # V13(età), V14, V15(tipo di alloggio)
plot(tree_cr)
text(tree_cr, pretty = 0)


pred_cr <- predict(tree_cr, newdata = cr_test, type = 'class')

conf_mat_cr <- table(Predicted = pred_cr, Actual = y_test_cr)
conf_mat_cr
#             Actual
# Predicted   1   2
#         1 177  70
#         2  19  34

tree_accuracy <- mean(pred_cr == y_test_cr)
tree_accuracy
#[1] 0.7033333

test_error_cr <- mean(pred_cr != y_test_cr)
test_error_cr
#[1] 0.2966667

recall_bad_credit <- conf_mat_cr[2,2]/(conf_mat_cr[1,2]+conf_mat_cr[2,2])
recall_bad_credit
# 0.3269

specificity <- conf_mat_cr[1,1]/(conf_mat_cr[1,1]+conf_mat_cr[2,1])
specificity
# 0.9031
```

The classification tree achieves an accuracy of about 70%.
However, it has difficulty identifying bad credit risks, showing a low recall for the minority class.

The problem are the 70 bad credits classified as good credit. 
In credit scoring, this is the worst possible error.
Let's consider "bad" (class 2) as a positive class.
The sensitivity is about 32.7%.
The tree intercepts only 32.7% of bad credits. Very low.
Specificity is about 90%, which is very high. The tree is good at intercepting good credits.

Let's see if techniques like Pruning, Bagging, Random Forest and Boosting can improve these values.


5. PRUNING

```{r}
set.seed(123)
cv <- cv.tree(tree_cr, FUN = prune.misclass)
plot(cv$size, cv$dev, type = 'b',
     xlab = 'Tree size',
     ylab = 'CV misclassification error',
     main = 'CV for classification tree (CR)')
best_size_cr <- cv$size[which.min(cv$dev)]
best_size_cr

tree_cr_pruned <- prune.misclass(tree_cr, best = best_size_cr)
plot(tree_cr_pruned)
text(tree_cr_pruned, pretty = 0)

pred_cr_pruned <- predict(tree_cr_pruned,
                               newdata = cr_test,
                               type = 'class')

conf_mat_cr_pruned <- table(Predicted = pred_cr_pruned,
                                 Actual    = y_test_cr)
conf_mat_cr_pruned
#                  Actual
#         Predicted   1   2
#                 1 178  70
#                 2  18  34

test_error_cr_pruned <- mean(pred_cr_pruned != y_test_cr)
test_error_cr_pruned
# 0.29 lievissimo miglioramento, quasi nullo.

recall_bad_credit_pruned <- conf_mat_cr_pruned[2,2]/(conf_mat_cr_pruned[1,2]+conf_mat_cr_pruned[2,2])
recall_bad_credit_pruned
```

Pruning slightly improves the model, but the improvement is limited.
The sensitivity (Recall for bad credit) is about 33%. Very low.


6. BAGGING
```{r}
set.seed(123)
positive_class <- levels(cr$Classe)[2] # "bad"
positive_class

library(tree)         # single regression tree
library(randomForest) # bagging / random forests

p <- ncol(cr) - 1  # number of predictors (excluding classe)

set.seed(1)
bag_cr <- randomForest(
  formula    = Classe ~ .,
  data       = cr_train,
  mtry       = p,       # all variables at each split -> bagging
  ntree      = 1000,     # number of trees in the ensemble
  importance = TRUE
)

bag_cr

pred_cr_bag <- predict(bag_cr, newdata = cr_test, type = 'class')
conf_mat_cr_bag <- table(Predicted = pred_cr_bag, Actual = y_test_cr)
conf_mat_cr_bag
#         Actual
#Predicted   1   2
#        1 169  56
#        2  27  48

tree_accuracy_bag <- mean(pred_cr_bag == y_test_cr)
tree_accuracy_bag
#[1] 0.7233333

test_error_cr_bag <- mean(pred_cr_bag != y_test_cr)
test_error_cr_bag
#[1] 0.2766667

recall_bad_credit_bag <- conf_mat_cr_bag[2,2]/(conf_mat_cr_bag[1,2]+conf_mat_cr_bag[2,2])
recall_bad_credit_bag
# 0.4615

specificity_bag <- conf_mat_cr_bag[1,1]/(conf_mat_cr_bag[1,1]+conf_mat_cr_bag[2,1])
specificity_bag
# 0.8622


importance(bag_cr)
varImpPlot(bag_cr,
           main = 'Variable importance (Bagging, cr data)')
```

Bagging improves the performance compared to the single decision tree.
It increases the ability to correctly identify bad credit risks by reducing model variance.

Recall(sensitivity) for bad is about 46.2%. 
Before with the single decision tree, it was 32.7%. Huge improvement.
Specificity is about 86%. Slightly worse than the tree, but still good. This model is catching more bad losing a bit on good.


7. RANDOM FOREST

```{r}
p_rf <- floor(sqrt(p))
p_rf

set.seed(2)
rf_cr <- randomForest(
  formula    = Classe ~ .,
  data       = cr_train,
  mtry       = p_rf,
  ntree      = 1000,
  importance = TRUE
)

rf_cr

pred_cr_rf <- predict(rf_cr, newdata = cr_test, type = 'class')
conf_mat_cr_rf <- table(Predicted = pred_cr_rf, Actual = y_test_cr)
conf_mat_cr_rf

tree_accuracy_rf <- mean(pred_cr_rf == y_test_cr)
tree_accuracy_rf
#[1] 0.6933

test_error_cr_rf <- mean(pred_cr_rf != y_test_cr)
test_error_cr_rf
#[1] 0.3067

plot(rf_cr,
     main = 'Random forest on CR data (OOB error vs number of trees)')
# Aumentando gli alberi si assesta a circa 21.86%

importance(rf_cr)
varImpPlot(rf_cr,
           main = 'Variable importance (Random forest, cr data)')

```

Random Forest does not outperform bagging in this case.


8. TUNING mtry

```{r}
m_test <- c(4, 6, 8, 10, 15)

rf_results <- c()

for (m in m_test) {
  
  set.seed(2)
  
  rf_model <- randomForest(
    Classe ~ .,
    data = cr_train,
    mtry = m,
    ntree = 1000
  )
  
  pred <- predict(rf_model, newdata = cr_test)
  
  rf_results <- c(rf_results, mean(pred == y_test_cr))
}

data.frame(mtry = m_test, accuracy = rf_results) # Tende all'accuracy del Bagging

```

Increasing the number of predictors considered at each split improves performance and makes the model closer to Bagging.


9. BOOSTING

```{r}
library(adabag)

set.seed(123)

boost_cr <- boosting(
  Classe ~ .,
  data    = cr_train,
  boos    = TRUE,
  mfinal  = 100,
  control = rpart.control(cp = 0.01,
                          minsplit = 3)
)

#------------------

pred_cr_boost <- predict(boost_cr, newdata = cr_test)
names(pred_cr_boost)

conf_mat_cr_boost <- table(Predicted = pred_cr_boost$class,
                          Actual    = cr_test$Classe)
conf_mat_cr_boost
#         Actual
# Predicted   1   2
#         1 157  51
#         2  39  53

accuracy_boost <- sum(diag(conf_mat_cr_boost)) / sum(conf_mat_cr_boost)
accuracy_boost
# 0.70

recall_bad_credit_boost <- conf_mat_cr_boost[2,2]/(conf_mat_cr_boost[1,2]+conf_mat_cr_boost[2,2])
recall_bad_credit_boost
# 0.5096

specificity_boost <- conf_mat_cr_boost[1,1]/(conf_mat_cr_boost[1,1]+conf_mat_cr_boost[2,1])
specificity_boost
# 0.80
```

Boosting manages to intercept half of the bad risks at the cost of lower specificity. 

10. MODEL COMPARISON AND CONCLUSION
```{r}
data.frame(
  Model = c("Tree","Bagging","Random Forest","Boosting"),
  Accuracy = c(tree_accuracy,
               tree_accuracy_bag,
               tree_accuracy_rf,
               accuracy_boost)
)
```

Tree-based methods were applied to classify credit risk.
The decision tree provides a baseline performance of around 70%.
Bagging improves performance and provides the best results overall in this analysis.
However, considering that intercepting bad credit is very important in credit scoring, we can say that Boosting is equally valid.

Random Forest does not outperform bagging, likely due to the limited number of predictors.

Overall, ensemble methods provide more stable predictions compared to a single decision tree.